---
title: "California Housing Prices Prediction - Data Visualization"
author: "Balaji Bharadwaj"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The housing market is a complex and ever-changing system. There are many factors that can
affect the price of a home, including location, size, condition, and amenities. 

In recent years,
machine learning algorithms have been increasingly used to predict housing prices.
 Machine learning algorithms are trained on large datasets of
historical data, and they use this data to learn patterns and relationships.

For this project, I will be using the dataset of housing prices from the San
Francisco Bay Area in order to build the models. Firstly, I will be examining and
visualizing the dataset to many observations regarding the same.

```{r}
## Loading the required libraries
library(ggplot2)
library(gridExtra)
library(dplyr)
library(reshape2)


## Loading the dataset
housing <- read.csv("C:/Users/Balaji Bharadwaj/Downloads/UB Subjects/Predictive Analytics/Project/housing.csv")
head(housing)
```

We can see that there are 9 continuous variables and one categorical variable (Ocean Proximity). 
Let us look at different charts and graphs to make observations of the dataset.

## Data Visualization

Let us first look at the counts of the variables, latitude and longitude.

```{r}
continuous_vars <- c('longitude', 'latitude')
housing_cont <- housing %>% select(all_of(continuous_vars))

plot1 <- ggplot(housing_cont, aes(x = latitude)) +
  geom_histogram(binwidth = 0.5, fill = 'blue', color = 'black') +
  xlab('Latitude') +
  ylab('Count') +
  theme_minimal()

plot2 <- ggplot(housing_cont, aes(x = longitude)) +
  geom_histogram(binwidth = 0.5, fill = 'blue', color = 'black') +
  xlab('Longitude') +
  ylab('Count') +
  theme_minimal()

grid.arrange(plot1, plot2, nrow = 1, ncol = 2, widths = c(3, 3))

```

We can see that a majority of the latitude entries lie at 34 degrees and at 37 degrees; and 
a majority of the longitude entries lie at -118 and -122 degrees. These regions correspond to
Los angeles and the San francisco bay area respectively.

Now let us look at the counts of the variables, Housing median age and Median Income.

```{r}
continuous_vars <- c('housing_median_age', 'median_income')
housing_cont <- housing %>% select(all_of(continuous_vars))

plot1 <- ggplot(housing_cont, aes(x = housing_median_age)) +
  geom_histogram(binwidth = 0.5, fill = 'blue', color = 'black') +
  xlab('housing_median_age') +
  ylab('Count') +
  theme_minimal()

plot2 <- ggplot(housing_cont, aes(x = median_income)) +
  geom_histogram(binwidth = 0.5, fill = 'blue', color = 'black') +
  xlab('median_income') +
  ylab('Count') +
  theme_minimal()

grid.arrange(plot1, plot2, nrow = 2, ncol = 1)

```

We can see a majority of housing is of age greater than 50 years. A majority of median income 
of families lie in the 200,000 - 400,000 range.

I will now generate scatter plots to see how each variable varies when compared to our target variable, 
median housing value.

```{r}

target <- housing$median_house_value


plots <- list()

for (i in seq_along(names(housing))) {
  col <- names(housing)[i]
  if (col != "median_house_value" && col != "ocean_proximity") {
    p <- ggplot(housing, aes_string(x = col, y = "median_house_value")) +
      geom_point(alpha = 0.5, size = 0.5) +
      xlab(col) +
      ylab("Median House Value") +
      theme_minimal()
    plots[[length(plots) + 1]] <- p
  }
}

grid.arrange(grobs = plots, nrow = 3, ncol = 3)

```

One of the main conclusions which we can draw from these scatterplots is that as median income increases, the 
median housing value also increases.

This can also be observed in the following heatmap. 

```{r}

numeric_housing <- housing[, sapply(housing, is.numeric)]
corr_matrix <- cor(numeric_housing)

# Melt the correlation matrix for visualization
melted_corr <- melt(corr_matrix)

# Create the heatmap
heatmap <- ggplot(melted_corr, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", 
                       midpoint = 0, limit = c(-1, 1), space = "Lab",
                       name="Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels
  labs(title = "Correlation Matrix", x = "", y = "") +  # Remove axis labels
  geom_text(aes(label = round(value, 2)), color = "black")  # Add annotations

# Display the plot
print(heatmap)

```

Apart from the obvious observations that:

1. population is correlated to the total rooms and households living in
 the block
 
 2. number of households being correlated to the total rooms
 
 we can see a strong correlation (0.69)
 between median income and medium housing value.


Let us now look at the categorical variable, Ocean_Promixity

```{r}
ocean_proximity_counts <- table(housing$ocean_proximity)

ocean_proximity_counts_df <- data.frame(Category = names(ocean_proximity_counts),
                                        Count = as.numeric(ocean_proximity_counts))

bar_plot <- ggplot(ocean_proximity_counts_df, aes(x = Category, y = Count)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(title = "Counts of Ocean Proximity", x = "Ocean Proximity", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


print(bar_plot)

```

We can see that a majority of houses are located less than 1 hour from the ocean followed by houses which
are inland. Houses that are present on a island make up a negligible amount of entries in the dataset.


